---
title: "Get gene-region associations for a group of GO terms"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: html_document
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 5,
    fig.height = 5,
    fig.align = "center"
)
options("width" = 100)
```

Let's first do a combination analysis with **rGREAT** and **simplifyEnrichment**. Let's say, you have
a list of genomic regions of interest (in the following example, we use a list of transcription factor
binding sites). You do a GO enrichment analysis with **rGREAT** and visualize the enrichment results
with **simplifyEnrichment**.


```{r, fig.width = 9, fig.height = 6}
library(rGREAT)
df = read.table(url("https://raw.githubusercontent.com/jokergoo/rGREAT_suppl/master/data/tb_encTfChipPkENCFF708LCH_A549_JUN_hg19.bed"))
# convert to a GRanges object
gr = GRanges(seqnames = df[, 1], ranges = IRanges(df[, 2], df[, 3]))
res = great(gr, "BP", "hg19")
tb = getEnrichmentTable(res)
head(tb)

library(simplifyEnrichment)
sig_go_ids = tb$id[tb$p_adjust < 0.001]
cl = simplifyGO(sig_go_ids)
head(cl)
table(cl$cluster)
```

The plot looks not bad. Next you may want to get the gene-region associations for significant GO terms. 
For example, the association for the first GO term:

```{r}
getRegionGeneAssociations(res, cl$id[1])
```

There are more than 600 significant GO terms, and perhaps you don't want to obtain such gene-region associations
for every one of them.

As **simplifyEnrichment** already clusters significant GO terms into a smaller number of groups, you may switch your
need of obtaining the associations from single GO terms to a group of GO terms where they show similar biological concept. In the new version of **rGREAT** (only available on GitHub currently), the second argument can be set as a vector of GO IDs. So, for example, to get the gene-region association for GO terms in the first cluster:

```{r}
getRegionGeneAssociations(res, cl$id[cl$cluster == 1])
```


## SessionInfo

```{r}
sessionInfo()
```

