---
title: "Work with other organisms"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: html_document
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center")
```

## Use TxDb packages

In the Bioconductor annotation ecosystem, there are **TxDb.\*** packages which
provide data for Gene Ontology gene sets. The **TxDb.\*** packages supported in **rGREAT** are:

```{r}
library(rGREAT)
rGREAT:::BIOC_ANNO_PKGS$txdb
```

To perform GREAT anlaysis with GO gene sets for other organisms, you can either
specify the genome version:

```{r, eval = FALSE}
great(gr, "GO:BP", "galGal6")
```

or with the full name of the corresponding TxDb package:

```{r, eval = FALSE}
great(gr, "GO:BP", "TxDb.Ggallus.UCSC.galGal6.refGene")
```

These two are internally the same.

## Use BioMart GO gene sets

You can specify a BioMart dataset (which corresponds to a specific organism), e.g.:

```{r, eval = FALSE}
# Giant panda
great(gr, "GO:BP", biomart_dataset = "amelanoleuca_gene_ensembl")
```

<p style="color:red">Make sure the genome version fit your data.</p>

A full list of supported BioMart datasets (organisms) as well as the genomo versions can be found with the function `BioMartGOGeneSets::supportedOrganisms()`.

## Use MSigDB gene sets

MSigDB contains gene sets only for human, but it can be extended to other organisms
by mapping to the homologues genes. The package [**msigdbr**](https://cran.r-project.org/web/packages/msigdbr/index.html) has
already mapped genes to many other organisms. A full list of supported organisms in **msigdbr** can be obtained by：

```{r}
library(msigdbr)
msigdbr_species()
```

To obtain gene sets for non-human organisms, e.g.:

```{r}
h_gene_sets = msigdbr(species = "chimpanzee", category = "H")
head(h_gene_sets)
```

If the organism you selected has a corresponding TxDb package available which provides TSS information, 
you need to make sure the gene sets use Entrez gene ID as the identifier (Most TxDb packages use Entrez ID
as primary ID, you can check the variable `rGREAT:::BIOC_ANNO_PKGS`).

```{r}
# convert to a list of gene sets
h_gene_sets = split(h_gene_sets$entrez_gene, h_gene_sets$gs_name)
h_gene_sets = lapply(h_gene_sets, as.character)  # just to make sure gene IDs are all in character.
h_gene_sets[1:2]
```

Now we can perform the local GREAT analysis.

```{r, eval = FALSE}
great(gr, h_gene_sets, "panTro6")
```

## Use NCBI genomes and gene sets

NCBI provides genome data for a huge number of organisms. It is quite easy to
obtain the GO gene sets for a specific organism with the **AnnotationHub**
package where NCBI is also the data source. Users please go to the
documentation of **AnnotationHub** to find out how to get GO gene sets with it.

It is relatively more difficult to obtain gene coordinates for a specific organism. 
Here the function `getGenomeDataFromNCBI()` accepts a
RefSeq accession number for a given genome assembly and returns information of the
genome as well as the genes.

The RefSeq accession number can be obtained from
https://www.ncbi.nlm.nih.gov/data-hub/genome/. Here we take dolphin as an
example. Its accession number is "GCF_011762595.1" and its genome page is at
https://www.ncbi.nlm.nih.gov/data-hub/genome/GCF_011762595.1/. We can get its genes with the command:

```{r, eval = .Platform$OS.type == "unix"}
genes = getGenomeDataFromNCBI("GCF_011762595.1", return_granges = TRUE)
genes
```

The chromosome length information is also included in `genes`:

```{r, eval = .Platform$OS.type == "unix"}
seqlengths(genes)
```

Now you can send `genes` to `great()` to perform the enrichment analysis.

```{r, eval = FALSE}
great(gr, your_go_gene_sets, tss_source = genes, ...)
```


The `GRanges` object can only be constructd if the genome is assembled on the
"chromosome-level". If it is not, e.g. on the scaffold or on the contig level.
`getGenomeDataFromNCBI()` returns the raw output which includes two data frames,
one for the genome, and one for the genes. Let's still take "GCF_011762595.1"
as an example, but set `return_granges` to `FALSE` (which is the default value in the function):

```{r, eval = .Platform$OS.type == "unix"}
lt = getGenomeDataFromNCBI("GCF_011762595.1", return_granges = FALSE)
names(lt)
```

The first several rows in the two data frames.

```{r, eval = .Platform$OS.type == "unix"}
head(lt$genome)
head(lt$gene)
```

In the original data from NCBI, genes are associated to contigs, and that is why
the first column in `lt$gene` contains RefSeq contig IDs. When the genome is
assembled on the chromosome level, the contigs are chromosomes. For some organisms,
there is no official chromosome name, then a specific contig ID from `lt$genome`
can be used to map to the contig names.

In the following code, we assume in users' input regions, the Genbank accession IDs
are used for contigs, then we need to construct a gene `GRanges` object which also takes
Genback accession IDs. This can be easily done as follows:

First we construct a mapping vector between Genbank IDs and RefSeq IDs:

```{r, eval = .Platform$OS.type == "unix"}
map = structure(lt$genome$genbankAccession, names = lt$genome$refseqAccession)
head(map)
```

Next apply `map` on the first column of `lt$gene` and then convert it into a `GRanges` object.
Just make sure in the ID convertion, `NA` may be procudes and need to be removed.

```{r, eval = .Platform$OS.type == "unix"}
genes = lt$gene
genes[, 1] = map[ genes[, 1] ]
genes = genes[!is.na(genes[, 1]), ]
genes = GRanges(seqnames = genes[, 1], ranges = IRanges(genes[, 2], genes[, 3]),
                strand = genes[, 4], gene_id = genes[, 5])
```

In `lt$genome` there is also contig length information. Simply construct it and set it to 
the `seqlengths` of `genes`.


```{r, eval = .Platform$OS.type == "unix"}
sl = structure(lt$genome$length, names = lt$genome$genbankAccession)
sl = sl[!is.na(names(sl))]
seqlengths(genes) = sl
genes
```

And similarly, `genes` can be directly sent to enrichment analysis.

```{r, eval = FALSE}
great(gr, your_go_gene_sets, tss_source = genes, ...)
```

Let's demonstrate how to obtain GO gene sets for a less-studied organism from
the `OrgDb` object. Still taking dolphin as an example, we can get its gene
coordinates with the `getGenomeDataFromNCBI()` function by specifying its
GenBank accession number. The `OrgDb` object for dolphin can be obtained with
the **AnnotationHub** package. First construct an `AnnotationHub` object:

```{r, eval = FALSE}
library(AnnotationHub)
ah = AnnotationHub()
```

Next we need to search dolphin on Annotation hub. It is sometimes not easy to
find the dataset from AnnotationHub as you want. In this example, we use query
words of "truncatus" and "OrgDb" (unfortunately, the word "dolphin" can not
find the correct file).

```{r, eval = FALSE}
query(ah, c("truncatus", "OrgDb"))
```

The returned message shows the ID of the OrgDb database for dolphin is
"AH112418". Then we can directly obtain the `OrgDb` object by:

```{r, eval = FALSE}
orgdb = ah[["AH112418"]]
```

With another helper function `getGeneSetsFromOrgDb()`, we can obtain the GO
gene sets for dolphin:

```{r, eval = FALSE}
go_gs = getGeneSetsFromOrgDb(orgdb)
great(gr, go_gs, tss_source = genes, ...)
```

## Use KEGG gene sets

KEGG supports a huge number of organisms. How to obtain the pathway gene sets
for an organism has been introduced in the vignette ["Work with other geneset
databases"](other-geneset-databases.html#kegg-pathways). Now the task is
to obtain the gene/TSS definitions of that organism. On the KEGG database, each
organism has a RefSeq accession ID associated and this ID can be found on the
webpage of the organism. For example, the web page of turkey is
https://www.genome.jp/kegg-bin/show_organism?org=mgp where the
RefSeq assembly access ID is "GCF_000146605.3". This ID can be used later
with `getGenomeDataFromNCBI()` to get the gene coordinates of that genome assembly.

**rGREAT** has a function `getKEGGGenome()` which automatically extracts the RefSeq
accession ID for an organism:

```{r}
getKEGGGenome("mgp")
```

Then, the following code performs GREAT analysis for any organism supported on KEGG, e.g. turkey:

```r
genes = getGenomeDataFromNCBI(getKEGGGenome("mgp"))
gene_sets = getKEGGPathways("mgp")

great(..., gene_sets, genes)
```

## Self-define TSS and gene sets

Since `great()` allows both self-defined TSS and gene sets, this means `great()` can be independent to organisms. 
Please refer to the vignette ["Analyze with local GREAT"](local-GREAT.html#manually-set-gene-sets-and-transcriptome-annotations) to
find out how to manuallly set both TSS and gene sets.


