---
title: "Work with other geneset databases"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_collapsed: false
    toc_float: true
vignette: >
  %\VignetteIndexEntry{4. Work with other geneset databases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center")
```


**rGREAT** has integrated Gene Ontology gene sets for all supported organisms and MSigDB gene sets for human. 
It is very easy to apply GREAT analysis on other geneset databases such as KEGG. THe basical use has been
explained in [the section "**Manually set gene sets**" in the vignette "**Analyze with local GREAT**"](local-GREAT.html#manually-set-gene-sets).
In this vignette, we will just show you several examples.


## KEGG pathways

[KEGG pathway database](https://www.genome.jp/kegg/pathway.html) is widely used for gene set enrichment analysis. To apply
GREAT anlaysis with KEGG pathways, we need to construct the corresponding gene sets.

KEGG provides its data via a REST API (https://rest.kegg.jp/). There are several commands that can be used to retrieve specific types of data.
Here we use the package [**KEGGREST**](https://bioconductor.org/packages/KEGGREST/) to obtain data directly from KEGG.
To get a list of genes involved in pathways, we use the "link" command which is implemented in the function `keggLink()`. 

```{r}
library(KEGGREST)
pathway2gene = keggLink("pathway", "hsa")
head(pathway2gene)
```

The returned object `pathway2gene` is a named vector, where the names are gene
Entrez IDs (with an additional `"hsa:"` prefix) and values are pathway IDs
(with an additional `"path:"` prefix). You can try to execute `keggLink("hsa","pathway")` 
to compare the results. The named vectors are not common for
downstream gene set analysis. Let's convert it to a list of genes, also remove the prefix.


```{r}
kegg_pathways = split(gsub("hsa:", "", names(pathway2gene)),
	                  gsub("path:", "", pathway2gene))
```

Now simply send `kegg_pathways` to `great()` to perform GREAT analysis.

```{r}
library(rGREAT)
gr = randomRegions(genome = "hg19")
great(gr, kegg_pathways, "hg19")
```

Here we only include the KEGG pathway IDs. The full name of pathways can be obtained by the following command:

```{r}
pn = keggList("pathway/hsa")
head(pn)
```

Please go the the documentation of **KEGGREST** or the main website of KEGG API (https://rest.kegg.jp/) to find out
commands for retrieving various datasets there.

<p style="color:red;">Please note, [KEGG pathways are only free to use for academic users, https://www.pathway.jp/en/licensing.html.</p>


## Reactome pathways

[Reactome](https://reactome.org/) is another popular pathway database for gene set enrichment analysis. 
In Bioconductor, there is a package [**reactome.db**](https://bioconductor.org/packages/reactome.db/) that
contains genes associated to every pathway in Reactome.

To get the Reactome pathway gene sets, simply use the object `reactomePATHID2EXTID`.

```{r}
library(reactome.db)
gs = as.list(reactomePATHID2EXTID)
great(gr, gs, "hg19")
```

The full pathway names are in the object `reactomePATHID2NAME`. You can convert the object to a normal list by `as.list(reactomePATHID2NAME)`.

## UniProt keywords gene sets

UniProt database provides a list of controlled vocabulary represented as keywords for genes or proteins (https://www.uniprot.org/keywords/). 
This is useful for summarizing gene functions in a compact way. The [**UniProtKeywords**](https://bioconductor.org/packages/UniProtKeywords/) package contains
gene sets associated to keywords.

**UniProtKeywords** support several organisms. Here we use human as an example.
In the function `load_keyword_genesets()`, you need to specify the taxon ID for the organism.

```{r}
library(UniProtKeywords)
gs = load_keyword_genesets(9606)
great(gr, gs, "hg19")
```


## Specific gene sets in .gmt format

Please refer to [the section "**Manually set gene sets**" in the vignette "**Analyze with local GREAT**"](local-GREAT.html#manually-set-gene-sets). 
There introduces a function `read_gmt()` which can import a `.gmt` file into a list of genes.

The tool [**Enrichr**](https://maayanlab.cloud/Enrichr/) provides a great resource of gene sets in `.gmt` format. See https://maayanlab.cloud/Enrichr/#libraries.


## Session info

```{r}
sessionInfo()
```
